From clusters to queries: exploiting uncertainty in the modularity landscape of complex networks

### 2018-04-09 article/article.tex
Added mann-whitney u test for query set significance.

\listfiles
\documentclass[manuscript, review, sigconf]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{color}

@@

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{images/search_space_3.png}
 \caption{
 Modularity search space of the \textit{E. coli} metabolic network.
 Distance between partitions is calculated using variation of information \cite{meilua2003comparing} and dimensionality reduction is performed using curvilinear component analysis \cite{demartines1997curvilinear}.
  The inset (top) demonstrates the landscape of the high modularity region. 
  Figure generated with the software of Good \textit{et al.}\cite{good2010performance}.
 }
 \label{fig:modular_search_space}
\end{figure}

@@

\subsection{Measuring the quality of relationships} \label{sec:expansion}
Given a query of vertices, the relatedness to other vertexes in a network is quantifiable by the fraction of times they are clustered with the query set, given the set of high quality partitions.
Formally, this can be expressed in terms of the \textit{expansion score} of a given vertex,
% \begin{equation} \label{eq:mu_score}
% \mu_i(S) = \frac{1}{|\mathcal{P}| |S - i|} \sum_{P \in \mathcal{P}} \sum_{j \in S - i} \delta(c^{P}_{i}, c^{P}_{j}),
% \end{equation}
\begin{equation} \label{eq:mu_score}
\mu_i(S) = \frac{1}{|\mathcal{P}|} \sum_{P \in \mathcal{P}} \delta(c^{P}_i, c^{P}_S),
\end{equation}
where $S$ denotes a query set, $P$ is a given partition in the space of all high quality partitions $\mathcal{P}$, $c^{P}_{i}$ indicates the partition vertex $i$ is contained in within partition $P$ and
$\delta(u, v)$ is the Dirac delta function that equals 1 if are the same cluster and $0$ otherwise, where $c^{P}_S$ is given by,
\begin{equation}
 c^{P}_S = \argmax_{C \in P} |C \cap S|,
\end{equation}
or the largest intersecting cluster with the query set, in partition $P$.
Using the largest intersecting community with $S$ is both computationally favourable for computing $mu_i$ for all $i$ in $V$ and allows the query set to contain potentially suprious query nodes.
Borrowing the terminology of fuzzy logic \cite{zimmermann2001fuzzy}, \textit{$\mu_i(S)$} can be considered a membership function and we can interpret this as a measure of the \textit{possibility} for which a given vertex belongs to the query set, given a set of partitions.
Where $\mu_i(S) = 1$ the partition space indicates that a given vertex is always partitioned in the same group as the query set.
In the case $\mu_i(S) = 0$ the vertex $i$ never appears in the same cluster as the largest query set.

The subtraction of the vertex from the query set ($S - i$) allows an unbiased score for vertices within the query set.
This allows us to define a vector, $\mu(S)$ for all vertices in the network, including those in $S$.

@@

\subsection{Synthetic networks}
In this section we test the method on a number benchmark networks constructed with a known, ground-truth community structure.
To evaluate how our method performs we demonstrate results against two benchmark networks the conventionally used LFR benchmark \cite{lfr} and a newer form of benchmark for community detection algorithms developed in previous studies \cite{cigram}.
Both these benchmarks allow for a configurable level of overlapping and mixing between communities.

\subsubsection{LFR benchmarks}
In this section we test the area under the ROC curve (AUC) scores for networks synthetically generated with the LFR benchmark.
The community distribution is defined with a powerlaw coefficient of $-1.0$ and the degree distribution is defined with a powerlaw coefficient of $-2.0$.
\textbf{Non-overlapping modules.} Figure \ref{fig:auc_no_overlap} represents tests on 100 networks with different randomly selected seed nodes.
As one would expect, the prediction of the method drops off steeply where communities are less defined above a mixing coefficient of 0.6.
Using larger seed sets also improves accuracy with some prediction of true communities being possible at extremely high levels of mixing.

@@

\subsubsection{CiGRAM benchmarks}

@@

 \item \textbf{Yeast protein-protein interaction network (Yeast PPI)} \cite{yeast_ppi}
 This dataset is a collection of recorded binary interactions between proteins collected with high-throughput yeast-2-hybrid assays.
 The metadata used are known, experimentally validated protein complexes from \cite{yeast_ppi_complexes}.
 The network contains 6222 nodes in the largest connected component, with 22,3868 edges.
 There are 409 experimentally validated protein complexes, 236 of which contain 3 or more nodes.
 The protein complexes are typically very small in terms of number of proteins with only 90\% of the complexes containing less than 10 proteins and only 2 complexes containing 50 or more proteins.

@@

>>>>

\listfiles
\documentclass[manuscript, proceedings]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{color}

@@

\begin{figure}
\centering
 \includegraphics[width=0.48\textwidth]{images/search_space_3.png}
 \caption{
 Modularity search space of an \textit{E. coli} metabolic network.
 Distance between partitions is calculated using variation of information \cite{meilua2003comparing} and dimensionality reduction is performed using curvilinear component analysis \cite{demartines1997curvilinear}.
  The inset (top) demonstrates the landscape of the high modularity region. 
  Figure generated with the software of Good \textit{et al.}\cite{good2010performance}.
 }
 \label{fig:modular_search_space}
\end{figure}

@@

\subsection{Measuring the quality of relationships} \label{sec:expansion}
Given a query of vertices, the relatedness to other vertexes in a network is quantifiable by the fraction of times they are clustered with the query set, given the set of high quality partitions.
Formally, this can be expressed in terms of the \textit{expansion score} of a given vertex,
% \begin{equation} \label{eq:mu_score}
% \mu_i(S) = \frac{1}{|\mathcal{P}| |S - i|} \sum_{P \in \mathcal{P}} \sum_{j \in S - i} \delta(c^{P}_{i}, c^{P}_{j}),
% \end{equation}
\begin{equation} \label{eq:mu_score}
\mu_i(S) = \frac{1}{|\mathcal{P}|} \sum_{P \in \mathcal{P}} \delta(c^{P}_i, c^{P}_S),
\end{equation}
where $S$ denotes a query set, $P$ is a given partition in the space of all high quality partitions $\mathcal{P}$, $c^{P}_{i}$ indicates the partition vertex $i$ is contained in within partition $P$ and
$\delta(u, v)$ is the Dirac delta function that equals 1 if are the same cluster and $0$ otherwise, where $c^{P}_S$ is given by,
\begin{equation}
 c^{P}_S = \argmax_{C \in P} |C \cap S|,
\end{equation}
or the largest intersecting cluster with the query set, in partition $P$.
Using the largest intersecting community with $S$ is both computationally favourable for computing $mu_i$ for all $i$ in $V$ and allows the query set to contain potentially spurious query nodes.
Where $\mu_i(S) = 1$ the partition space indicates that a given vertex is always partitioned in the same group as the query set.
In the case $\mu_i(S) = 0$ the vertex $i$ never appears in the same cluster as the largest query set.
We define $\mu_i(S)$ for all vertices in the network, including those in $S$.
In many cases, a query set will not be fully represented by the partitions found within the network, this is further elaborated in the following section.

\subsection{Statistically significant query sets}
Naturally, not every label set will be of high quality or represented within the community structure of a network.
This is established further in section \ref{sec:go_labels} which tests gene ontology terms (GO) on protein interaction networks.
In this case, all the labels relate to the real function of the vertices, however, GO is a hierarchical system and many terms are extremely general.
Furthermore, even for specific terms the network datasets will often not be well represented.
A common approach with community detection and clustering (used for example in \cite{dekkers2013}) is to detect communities and find clusters containing significant GO terms.

In this work, we would wish to use the community partition space, described above, to find highly related sets of vertices.
For this we take the distribution, from Equation \ref{eq:mu_score} for a given query set.
For a query set to be relevant it should be significantly more related than one would expect from a set of nodes selected at random within a network.

Formally, given the partition space $\mathcal{P}$ and a query set $S$ we consider the distribution of quality scores as $\mu_i(S - i)$, for all $i$ in $S$.
In order for $S$ to be a high quality query set, its value of internal vertices should be significantly higher than for external vertices.
Formally, we assert that the distribution $\mu_i(S - i)$ should reject the null hypothesis that the mean is not significantly greater than $\mu_j(S)$ where $j \in S - V$, the set of vertices excluding those in the query set.
For the purposes of this work we use the non-parametric Mann–Whitney $U$ test and reject the null hypothesis where $p < 0.01$.
In section \ref{sec:go_labels} this is used to exclude less relevant GO terms.

@@

\subsection{Synthetic networks}
In this section we test the method on a number benchmark networks constructed with a known, ground-truth community structure.
To evaluate how our method performs we demonstrate results against two benchmark networks the conventionally used LFR benchmark \cite{lfr} for community detection algorithms.
We test the area under the ROC curve (AUC) scores for networks synthetically generated with the LFR benchmark.
The community distribution is defined with a powerlaw coefficient of $-1.0$ and the degree distribution is defined with a powerlaw coefficient of $-2.0$.
\textbf{Non-overlapping modules.} Figure \ref{fig:auc_no_overlap} represents tests on 100 networks with different randomly selected seed nodes.
As one would expect, the prediction of the method drops off steeply where communities are less defined above a mixing coefficient of 0.6.
Using larger seed sets also improves accuracy with some prediction of true communities being possible at extremely high levels of mixing.

@@

@@

 \item \textbf{Yeast protein-protein interaction network (Yeast PPI)} \cite{yeast_ppi}
 This dataset is a collection of recorded binary interactions between proteins collected with high-throughput yeast-2-hybrid assays.
 The metadata used are known, experimentally validated protein complexes from \cite{yeast_ppi_complexes}.
 The network contains 6222 nodes in the largest connected component, with 22,3868 edges.
 There are 409 experimentally validated protein complexes, 236 of which contain 3 or more nodes.
 The protein complexes are typically very small in terms of number of proteins with  90\% of the complexes containing less than 10 proteins and only 2 complexes containing 50 or more proteins.

@@

\subsection{Gene ontology data}
\label{sec:go_labels}
